### 2020-11-10 latex/sections/methods.tex
iteration

In this chapter we describe our implementation of the game Breakthrough and the methods for training a neural network to play the game Breakthrough, described in Chapter \ref{cha:background}. The training process uses MCTS described in Section \ref{sec:mcts} to guide its training. We outline the architecture used in the neural network.

The training algorithm is a reinforcement learning self-play algorithm based on previous work by DeepMind (CITE ME).

\section{Implementation}

In this section we will take a brief look at the implementation of the game Breakthrough, how we model its state representation, and the pseudocode for training.

@@

In order for a neural network to be able to use input, that input needs to be numeric. We model the board in gamethrough with a single $NxM$ array with three values $\{-1,0,1\}$. Where each $x,y$ index on the board represents what value is on the board at the time, $-1$ represnts a white piece, $0$ represents an empty square, and $1$ represents a black piece.

A action vector is coupled with each state, the action vector is a $3$-dimensional matrix where the lengths of the dimensions are $row x columns x 6$ and each index of the vector $X,Y,Z$ represents an action moving from cell $x,y$ moving to the direction $z$. The directions are as follows $0$ represents moving upwards and to the left diagonally on the board, $1$ upwards, and $2$ upwards and to the right diagonally, $4$ downwards and to the left diagonally, $5$ downwards, and lastly $6$ downwards and to the right diagonally.

\subsection{Selfplay}

@@

The neural network architecture we opted to use was a single convolutional layer with a ReLU activation function, followed by 5 residual layers each containing two layers of convolution, batch normalization and a ReLu activation. Lastly a split policy/value head, where the output of the last residual layer is split into two different outputs. The policy head are two layers: first a final convolutional layer, and last a fully connected layer with a \textit{log softmax} layer. The value head, consists of three layers: first a convolutional layer, then a fully connected layer, and lastly a fully connected layer with a \textit{tanh} activation function. The Figure \ref{fig:nnarch}. depicts the architecture.

@@

It is important to understand why we use convolutional layers when dealing with board games as convolution is often associated with image. This is because a convolutional layer is focused on merging multiple input parameters to a single neuron, making it an input parameter for the next layer representing the locality around the center point of the original input. Playing the game of Breakthrough, a piece in only able to capture pieces in its immediate vicinity. This lead to the selection of a convolutional kernel with the size of $3x3$ and a stride of $1$.

The knowledge of each piece moves isn't enough to represent the whole gamestate within the neural network. This is why we use multiple residual layers with convolution. These residual layers stack convolutions on each other making the final convolution represent the locality of all the other localities, allowing the neural network to have a representation of the whole gamestate in its parameters.

@@

The residual layers as a concept remember the output of the previous layer and add the results of the residual layer and the previous layer together. On a higher level this leads to the internal representation of the neural network to maintain the whole game boards in its representation, s.t. two areas that are far away from each other maintain the same level of locality as two that are close.

@@

The policy head of the neural network returns a vector of the size of the action space $w * h * 6$ where $w$ is the width of the board $h$ is the height of the board and $6$ represents the six cardinal direction pawns can move (straight, and diagonally both left and right for the first player, and backwards for the second player). The vector is then masked s.t. the values representing moves that can not be played on the board are given the value of $0$.

@@

@@

To train the neural network to play Breakthrough we initialize two neural networks $N_1$ and $N_2$ 
with random weights. Then we let $N_1$ play against itself using MCTS to direct its training. 
While playing against itself the neural network gathers data for it to train with.

The data collected is $(\pi, \tau, p, v)$, where $\pi$ is the action probabilities provided by 
MCTS for $N$ iterations, $\tau$ is the reward for the whole episode, $1$ if white wins,
$-1$ if black wins, $p$ is the policy vector
from the neural network, and $v$ is the predicted reward of the game by the neural network.

This data collection trains the neural network in such a way that it is a function approximation for a single state to return the values that would have been returned from applying MCTS for $N$ iterations.

@@

The data collected is backpropagated through the NN moving the weights to the direction of this 
loss function $l = (p * \pi) + (\tau - v)^2$ for each state the NN encountered during self-play.

@@

In this section we describe the

@@

Once the neural network has learned to play the game of Breakthrough, we examine its internal state w.r.t HLC's that we understand. The first examined HLC is numbers advantage, that is, the number of pieces that a player has over his opponent. The higher level idea to a human player would be that they are in a better position since they have more pieces.

To examine the higher level component we take a look at the internal state of the neural network itself. To do this we take a state, we run the state through the neural network, and while the state is propagating through the network we select a layer to split the network, we call that layer $l_{split}$. At that point in time the state is a mutated vector $l_{split}(l_{split-1}(...l_0(state)))$ of the initial state. We save that vector as a point in the N-dimensional space that it represents, and once we've gathered enough points in that N-dimensional space we're able to train a linear classifier using the HLC as a label. We train a Stochastic Gradient Descent classifier, from the SKLearn library, to construct a hyperplane that splits the space into two binary states, one that contains the HLC and another that doesn't. 
\begin{figure}[]
    \centering
    \includegraphics[width=0.7\textwidth]{graphics/linear_separation}
    \caption{Trained linear classifier on a 2-dimensional space}
    \label{fig:scattersplit}
\end{figure}
An example of how this would look with only two dimensions is shown in Figure \ref{fig:scattersplit} Once we've successfully trained this linear classifier we can run another state $s$ through the neural network. Once the prediction has finished, we view the selected action $a$ of $s$ from the policy vector $p$ from the neural network. We view the loss of that selected action $a$. Then we apply the backpropagation algorithm up to that same layer $l_{split}$ and evaluate the gradient there. If that gradient moves to the direction of the HLC we say that the state includes the HLC, and doesn't if it move away from the HLC.
\begin{figure}[]
    \centering
    \includegraphics[width=0.7\textwidth]{graphics/linear_separation_with_direction}
    \caption{Trained linear classifier on a 2-dimensional space with arrow representing gradient}
    \label{fig:scattersplitarrow}
\end{figure}
We show an example in Figure \ref{fig:scattersplitarrow}. where the arrow in the image represents the direction the gradient is moving, if it moves toward the HLC the state $s$ includes the HLC otherwise it does not. Using this method we evaluate the internal representation of the state within the neural network, and are able to see whether it recognizes the HLC.

>>>>

In this chapter, we describe our implementation of the game Breakthrough and the methods for training a neural network to play the game Breakthrough, described in Chapter \ref{cha:background}. The training process uses MCTS described in Section \ref{sec:mcts} to guide its training. We outline the architecture used in the neural network.

The training algorithm is a reinforcement learning self-play algorithm based on previous work by DeepMind~\cite{silver:alphagozero}.

\section{Game Implementation}

In this section, we will take a brief look at the implementation of the game Breakthrough, how we model its state representation and the pseudocode for training.

@@

For a neural network to be able to use input, that input needs to be numeric. We model the board in gamethrough with a single $NxM$ array with three values $\{-1,0,1\}$. Where each $x,y$ index on the board represents what value is on the board at the time, $-1$ represents a white piece, $0$ an empty square, and $1$ a black piece.

An action vector is coupled with each state, the action vector is a $3$-dimensional matrix where the lengths of the dimensions are $N * M * 6$ and each index of the matrix represents an action moving from cell $x,y$ moving to the direction $z$. The directions are as follows $0$ represents moving upwards and to the left diagonally on the board, $1$ upwards, and $2$ upwards and to the right diagonally, $4$ downwards and to the left diagonally, $5$ downwards, and lastly $6$ downwards and to the right diagonally.

@@

The neural network architecture we opted to use was a single convolutional layer with a ReLU activation function, followed by 5 residual layers each containing two layers of convolution, batch normalization, and a ReLu activation. Lastly, a split policy/value head, where the output of the last residual layer is split into two different outputs. The policy head are two layers: first a final convolutional layer, and then a fully connected layer with a \textit{log softmax} layer. The value head consists of three layers: first a convolutional layer, then a fully connected layer, and lastly a fully connected layer with a \textit{tanh} activation function. The Figure \ref{fig:nnarch}. depicts the architecture.

@@

It is important to understand why we use convolutional layers when dealing with board games as convolution is often associated with an image. This is because a convolutional layer is focused on merging multiple input parameters to a single neuron, making it an input parameter for the next layer representing the locality around the center point of the original input. Playing the game of Breakthrough, a piece is only able to capture pieces in its immediate vicinity. This lead to the selection of a convolutional kernel with a size of $3x3$ and a stride of $1$.

The knowledge of each piece moves isn't enough to represent the whole game state within the neural network. This is why we use multiple residual layers with convolution. These residual layers stack convolutions on each other making the final convolution represent the locality of all the other localities, allowing the neural network to have a representation of the whole game state in its parameters.

@@

The residual layers as a concept remember the output of the previous layer and add the results of the residual layer and the previous layer together. On a higher level, this leads to the internal representation of the neural network to maintain the whole game boards in its representation, s.t. two areas that are far away from each other maintain the same level of locality as two that are close.

@@

The policy head of the neural network returns a vector of the size of the action space $w * h * 6$ where $w$ is the width of the board $h$ is the height of the board and $6$ represents the six cardinal direction pawns can move (straight, and diagonally both left and right for the first player, and backward for the second player). The vector is then masked s.t. the values representing moves that can not be played on the board are given the value of $0$.

@@

\subsection{Implementation}

This subsection focuses on the implementation of the various functions required to train the neural network to play breakthrough. The description is mainly through pseudo code, and the code is awailable on \href{github.com/sigurdurhelga/msc-cheis}{Github} for examination.

@@

\begin{algorithm}[t]
\caption{Neural network selfplay pseudocode}
\label{alg:selfplay}
\begin{algorithmic}[1]
\STATE \textbf{neural network} nn1 = randomizedInitialNN()
\STATE \textbf{neural network} nn2 = randomizedInitialNN()
\WHILE{\TRUE}
    \STATE{dataset = \textbf{generate\_dataset}(nn1)}
    \STATE{nn1.train\_on\_examples(dataset)}
    \STATE{win\_rate = \textbf{compete}(nn1,nn2)}
    \IF{win\_rate > $0.5$}
        \STATE{nn2 = nn1.copy()}
    \ELSE
        \STATE{nn1 = nn2.copy()}
    \ENDIF
    \STATE \textbf{save}(nn1)
\ENDWHILE
\end{algorithmic}
\end{algorithm}

To train the neural network to play Breakthrough we initialize two neural networks nn1 and nn2 with random weights. Then we let nn1 play against itself using MCTS with PUCT to select moves. We collect data from this self-play to train on. The data collected is $(s, \pi, \tau, p, v)$, $s$ is the state, $\pi$ is the action probability vector provided by MCTS with PUCT, $\tau$ the end reward for the episode $1$ if white wins, $-1$ if black wins. $p$ is the action probability vector predicted by the neural network, and $v$ is the predicted reward of the game by the neural network.

The pseudocode can be seen in Algorithm~\ref{alg:selfplay}.

\subsection{Compete}

\begin{algorithm}[t]
\caption{Neural network compete pseudocode}
\label{alg:competition}
\begin{algorithmic}[1]
\STATE{\textbf{Input:} neural network nn1, neural network nn2}
\STATE{\textbf{Output:} win rate for white player}
\STATE{white\_wins $= 0$}
\STATE{game = \textbf{initial\_breakthrough}()}
\FOR{1 \TO 100}
    \WHILE{\textbf{not} game.is\_termina()}
        \STATE{nn1.make\_move()}
        \STATE{nn2.make\_move()}
    \ENDWHILE
    \IF{game.white\_wins()}
        \STATE{white\_wins++}
    \ENDIF
\ENDFOR
\STATE{\textbf{return} white\_wins/100}
\end{algorithmic}
\end{algorithm}

The compete function differs from the self-play function in such a way that we select the moves by a single pass through the neural network thereby only evaluating the neural networks ability to predict best actions. The pseudocode is available in Algorithm~\ref{alg:competition}.

@@

The data collected is backpropagated through the NN moving the weights to the direction of this loss function $l = (p * \pi) + (\tau - v)^2$ for each state the NN encountered during self-play. Importantly the variable $p$ has masked illegal actions to $0$ to direct the NN to not learn on illegal moves.

@@

In this section we describe the process of trainin a Concept Activation Vector in order to linearly separate each state into points in a space with the concept and ones without the concept.

@@

Once the neural network has learned to play the game of Breakthrough, we examine its internal state w.r.t HLC's that we understand. The first examined HLC is numbers advantage, that is, the number of pieces that a player has over his opponent. The higher-level idea to a human player would be that they are in a better position since they have more pieces.

To examine the higher-level component we take a look at the internal state of the neural network itself. To do this we take a state, we run the state through the neural network, and while the state is propagating through the network we select a layer to split the network, we call that layer $l_{split}$. At that point in time the state is a mutated vector $l_{split}(l_{split-1}(...l_0(state)))$ of the initial state. We save that vector as a point in the N-dimensional space that it represents, and once we've gathered enough points in that N-dimensional space we're able to train a linear classifier using the HLC as a label. We train a Stochastic Gradient Descent classifier, from the SKLearn library, to construct a hyperplane that splits the space into two binary states, one that contains the HLC and another that doesn't.

\begin{figure}[]
    \centering
    \includegraphics[width=0.7\textwidth]{graphics/linear_separation}
    \caption{Trained linear classifier on a 2-dimensional space}
    \label{fig:scattersplit}
\end{figure}

An example of how this would look with only two dimensions is shown in Figure \ref{fig:scattersplit} Once we've successfully trained this linear classifier we can run another state $s$ through the neural network. Once the prediction has finished, we view the selected action $a$ of $s$ from the policy vector $p$ from the neural network. We view the loss of that selected action $a$. Then we apply the backpropagation algorithm up to that same layer $l_{split}$ and evaluate the gradient there. If that gradient moves in the direction of the HLC we say that the state includes the HLC, and doesn't if it moves away from the HLC.

\begin{figure}[]
    \centering
    \includegraphics[width=0.7\textwidth]{graphics/linear_separation_with_direction}
    \caption{Trained linear classifier on a 2-dimensional space with arrow representing gradient}
    \label{fig:scattersplitarrow}
\end{figure}

We show an example in Figure \ref{fig:scattersplitarrow}. where the arrow in the image represents the direction the gradient is moving, if it moves toward the HLC the state $s$ includes the HLC otherwise it does not. Using this method we evaluate the internal representation of the state within the neural network and can see whether it recognizes the HLC.

### 2020-11-10 latex/sections/results.tex
iteration

\ifdraft{In this section you discuss any issues that came up while developing
the system.  If you found something particularly interesting,
difficult, or an important learning experience, put it here.  This is
also a good place to put additional figures and data.}
hello world

>>>>

\label{cha:results}

\ifdraft{In this section you discuss any issues that came up while developing
    the system.  If you found something particularly interesting,
    difficult, or an important learning experience, put it here.  This is
    also a good place to put additional figures and data.}

In this section we discuss the results of considering HLC within the game Breakthrough, using a Concept Activation vector for evaluating how the neural network recognizes the HLC's. We first evaluate the neural network against various agents to validate that it has learned how to play the game. Secondly, we take a look at the changes in the emphasis of the neural network during training. The main point of interest there being whether the neural network notices simple HLCs early then stops considering them as the network improves.

\section{Evolution of the neural networks win rate}

\begin{figure}
    \begin{small}
        \begin{center}
            \includegraphics[width=0.7\textwidth]{graphics/winratevsrandom.png}
        \end{center}
        \caption{Winrate vs random agent}
        \label{fig:winratevsrandom}
    \end{small}
\end{figure}

\begin{figure}
    \begin{small}
        \begin{center}
            \includegraphics[width=0.7\textwidth]{graphics/winratevsmcts.png}
        \end{center}
        \caption{Winrate vs monte carlo tree search agent}
        \label{fig:winratevsmcts}
    \end{small}
\end{figure}

\begin{figure}
    \begin{small}
        \begin{center}
            \includegraphics[width=0.7\textwidth]{graphics/winratevsneuralnetwork.png}
        \end{center}
        \caption{Winrate vs neural network trained for 150 generations}
        \label{fig:winratevsneuralnetwork}
    \end{small}
\end{figure}

Examining the neural network is only interesting if the neural network can effectively play the game. We first examine the history neural network playing against an agent that only takes random moves, on every $10$ generations. The win rate over generations can be seen in Figure \ref{fig:winratevsrandom}. It is clear that very early on in the training process, almost immediately after generation $10$ our agent performs nearly perfectly against the random agent. This implies that the agent does indeed learn some strategy in the game. The next agent we tried the neural network against, was an agent that does regular Monte-Carlo tree search on the game space with $100$ iterations of rollout for each move. A graph showing the win rate over generations can be seen in Figure \ref{fig:winratevsmcts}. Again, quickly our agent learns some strategy and can win often, but doesn't achieve a perfect win rate after $300$ generations. The last agent we tested against was the neural network itself, although only trained to $150$ generations. A graph showing the win rate can be seen in Figure \ref{fig:winratevsneuralnetwork}. As expected the agent performs poorly until it has trained for more than $150$ generations.

These results imply that our agent certainly understands how to play the game of breakthrough to a certain level. However, it should be stated that as we the creators attempt to compete against the agent it generally loses in a way that we would hope the agent should be able to play against. It is likely that as we train the neural network more it will become better.

\section{Evaluating the improvements of the neural network over generations}

To test which HLC the neural network places its emphasis on during training we trained a neural network for $300$ generations, taking snapshots of the network every $10$ generations. Then we had the neural network play against itself for $100$ games collecting the states it encountered during play. These states were then examined by a concept activation vector representing these HLCs.

We test four concepts \textit{Numbers Advantage}, \textit{Aggressiveness}, \textit{Unity}, and \textit{Lorentz-Horey}.

\subsection{Description of concepts}

\begin{figure}
    \begin{small}
        \begin{center}
            \includegraphics[width=0.95\textwidth]{graphics/dist_number_adv}
        \end{center}
    \end{small}
    \caption{Distribution of piece amount difference}
    \label{fig:dist_number_adv}
\end{figure}

The first concept we examine is Numbers Advantage which conveys the idea of having more pawns than your opponent. When faced with a board it is difficult to argue for whether having a single pawn up on your opponent constitues as Numbers Advantage or three. In order to an appropriate value for this case we used the neural network that had been trained for $300$ generations to generate $20000$ unique states. The distribution of the difference of pawns for each player can be seen in Figure~\ref{fig:dist_numbers_adv}. From examining the distribution and examining samples we selected the breakpoint of $2$, meaning that if you have $2$ pieces on your opponent you are in a state that has this concept. From the $20000$ states only $5.5\%$ of states have this property.

\begin{figure}
    \begin{small}
        \begin{center}
            \includegraphics[width=0.95\textwidth]{graphics/dist_aggressiveness}
        \end{center}
        \caption{Distribution of difference of furthest pawns}
        \label{fig:dist_aggressiveness}
    \end{small}
\end{figure}

The next concept is Aggressiveness, which describes how much closer your furthest piece is to the opponents edge than your opponents furthest piece to your edge. This is then the minimum amount of how many moves it will take you to win the game. A distribution of these values can be seen in Figure~\ref{fig:dist_aggressiveness}. We examined the distribution, and sampled states from various breaking points and decided on the value of $\geq 1$. That is if the value of a state's furthest piece difference is greater than or equal to $1$ that state has the concept of aggressiveness.

\begin{figure}
    \begin{small}
        \begin{center}
            \includegraphics[width=0.95\textwidth]{graphics/dist_unity}
        \end{center}
        \caption{Difference of distance from center}
        \label{fig:dist_unity}
    \end{small}
\end{figure}

\subsection{Numbers Advantage}

Firstly there is the numbers advantage HLC, where the number of pawns the player has minus the number of pawns the opponent had, the breakpoint we selected was $2$ meaning that if you have $2$ more pawns then your opponent, you're in a position where a HLC called numbers advantage is present.

\begin{figure}[]
    \centering
    \includegraphics[width=0.7\textwidth]{graphics/number_pawns_trend}
    \caption{Percentage of selected states containing the HLC numbers advantage}
    \label{fig:numberadvantage}
\end{figure}

The Figure \ref{fig:numberadvantage}. shows that throughout training the numbers advantage HLC is only ever a slight factor in the selection of states and we can say that the neural network doesn't consider number advantage in its selection process.

The second HLC we examined was aggressiveness, which is a state in which your most advanced pawn is $2$ or more squares further than the opponents most advanced pawn.

\begin{figure}[]
    \centering
    \includegraphics[width=0.7\textwidth]{graphics/most_advanced_trend.png}
    \caption{Percentage of selected states containing the HLC aggressiveness}
    \label{fig:aggressiveness}
\end{figure}

From Figure \ref{fig:aggressiveness}. we can see that the aggressiveness HLC is a growing factor over as the neural network is trained. Generally, when your opponent is not skilled this strategy is considered a good one.

The third HLC that we examined was unity, the unity HLC represents the absolute average distance of your pawns from the center row of your pawns. This HLC is calculated as the row of your furthest pawn from the starting row $r_{far}$, the row of your nearest pawn from the starting row $r_{near}$. Finding the middle row is then $\frac{r_{far} + r_{near}}{2}$, we then take the absolute of the average distance from the pawns to that row. To find the point at which this value relates to a state with the HLC unity, we sampled a myriad of states and decided on the value of $0.35$. The value of $0.35$ generally allows your states to have two rows that have the majority of the pawns and one or two pawns one row away from the group.

\begin{figure}[]
    \centering
    \includegraphics[width=0.7\textwidth]{graphics/unity_trend.png}
    \caption{Percentage of selected states containing the HLC unity}
    \label{fig:unity}
\end{figure}

The literature of breakthrough implies that being patient and waiting until your opponent makes a mistake is generally the preferred strategy for winning. the concept activation vector for Unity is not often triggered and is relatively stable at $50\%$ througout training.

\begin{table}[]
    \centering
    \begin{tabular}{|c|c|c|c|c|c|}
        \hline
        5  & 15 & 5  & 5  & 15 & 5  \\\hline
        2  & 3  & 3  & 3  & 3  & 2  \\\hline
        4  & 6  & 6  & 6  & 6  & 4  \\\hline
        7  & 10 & 10 & 10 & 10 & 7  \\\hline
        11 & 15 & 15 & 15 & 15 & 11 \\\hline
        21 & 21 & 21 & 21 & 21 & 21 \\\hline
    \end{tabular}
    \caption{Lorentz-Horey cell values, from white players point of view}
    \label{table:lorentzcell}
\end{table}

\begin{figure}
    \begin{small}
        \begin{center}
            \includegraphics[width=0.7\textwidth]{graphics/lorentz_horey_distribution.png}
        \end{center}
        \caption{Distribution of Lorentz Horey values}
        \label{fig:lorentzdistribution}
    \end{small}
\end{figure}

\begin{figure}
    \begin{small}
        \begin{center}
            \includegraphics[width=0.95\textwidth]{graphics/lorentz_horey_trend.png}
        \end{center}
        \caption{Percentage of selected states containing the HLC Lorentz-Horey}
        \label{fig:lorentzheuristic}
    \end{small}
\end{figure}

The last heuristic we examined was the Lorentz-Horey heuristic. This heuristic is a popular heuristic in breakthrough defined in the paper by Lorentz \& Horey~\cite{lorentz:heuristic}. For this heuristic, we give each cell on the board a point value. For a given player the heuristic is then the sum of the cell values where they have pawns. The cell values can be seen in Table \ref{table:lorentzcell}. the matrix shown is flipped in the black player's point of view. The values are selected in such a way that as your pawn approach the opponent's side their value increases, having pieces on the edge isn't optimal, as such a pawn will only be able to capture in one direction and can not escape capture as easily. To select the breakpoint for Lorentz-Horey heuristic we generated 20000 states and considered the values, the resulting distribution can be seen in Figure \ref{fig:lorentzdistribution}. This distribution leads us to select the value of $5$, as a breakpoint indicating that a state has the HLC of Lorentz-Horey. The evolution of the neural network using this HLC for selection can be seen in Figure \ref{fig:lorentzheuristic}

Examining Figure~\ref{fig:lorentzheuristic}, we can see that the trend of using this HLC increases over time. As this heuristic is the one most often cited as a successful strategy in breakthrough this result is promising.

\section{Examining state properties}

To be able to view the results from the concept activation vectors it is important to know how many states we expect to have a concept so that we can know whether the model is only guessing at random or not.
